---
title: "Default Predictors"
author: "Aidan Astridge"
date: 2023-04-18
subtitle: "April 2005 — September 2005 Taiwanese Consumer Credit Payments"
bibliography: references.bib
---

## Insight

Based on credit data, a higher credit balance and a higher payment made towards the previous month's bill are associated with a lower chance of defaulting on a credit account in the next month.

Compared to the average borrower of Female, Single, and University educated: Male borrowers increased defaulting chances, Married borrowers increased defaulting chances, and graduates that considered themselves "Other" decreased defaulting chances.

Delays in payment in September, July, and April are significant in predicting default for October. Surprisingly, August predicts the opposite.

## About the Data

The [data set](https://archive.ics.uci.edu/ml/datasets/default+of+credit+card+clients) was acquired from a research that focused on predicting the probability of default payments among customers in Taiwan using six data mining methods. Since the actual probability of default was unknown, the research introduced a new method called the "Sorting Smoothing Method" to estimate probability of default.

They then used the estimated probability as the response variable and the predictive probability of default as the independent variable in a simple linear regression. What they found was that an artificial neural network method had the highest coefficient of determination, with a regression intercept close to zero and a regression coefficient close to one. As a result, the artificial neural network method is the one they chose.

[@yeh2009]

> The data-set originally had 30,000 values; 70% of the data was taken out by filtering out below 5% and above 95% from a majority of the continuous variables. However, I've elected to keep the payment delays per month as it takes out to many variables

#### Variables

```{r include=FALSE}
library(tidyverse)
library(lubridate)
library(scales)
library(plotly)
library(ggthemes)
library(knitr)
library(kableExtra)
options(digits = 4)
options(scipen = 999)

cc <- read.csv("/Users/aidanastridge/Documents/default of credit card clients.csv")

#Data Wrangling

cc <- cc %>% 
  mutate(SEX = recode(SEX, `1` = "Male", `2` = "Female"), 
         EDUCATION = recode(EDUCATION, `1` = "Graduate", `2` = "University", `3` = "High School", `4` = "Others", `5` = "Others", `6` = "Others", `0` = "Others"),
         MARRIAGE = recode(MARRIAGE, `1` = "Married", `2` =  "Single", `3` = "Others", `0` = "Others"))

cc <- cc %>%
  rename(Balance = LIMIT_BAL,
         Sex = SEX,
         Education = EDUCATION,
         Marriage = MARRIAGE,
         Age = AGE,
         Default = default.payment.next.month)

status <- c("PAY_0","PAY_2","PAY_3","PAY_4","PAY_5","PAY_6")
bill <- c("BILL_AMT1","BILL_AMT2","BILL_AMT3","BILL_AMT4","BILL_AMT5","BILL_AMT6")
pay <- c("PAY_AMT1","PAY_AMT2","PAY_AMT3","PAY_AMT4","PAY_AMT5","PAY_AMT6")

cc <- cc %>% rename_at(vars(status), ~ paste(rev(seq(ym("2005-4"), ym("2005-9"), by = "months")), "Status"))
cc <- cc %>% rename_at(vars(bill), ~ paste(rev(seq(ym("2005-4"), ym("2005-9"), by = "months")), "Bill"))
cc <- cc %>% rename_at(vars(pay), ~ paste(rev(seq(ym("2005-4"), ym("2005-9"), by = "months")), "Payment"))

cc<-cc %>%
  filter(Balance > quantile(Balance,c(.05)) & Balance < quantile(Balance,c(.95))
         & Age > quantile(Age,c(.05)) & Age < quantile(Age,c(.95))
         & cc$`2005-09-01 Bill` > quantile (cc$`2005-09-01 Bill`,c(.05)) & cc$`2005-09-01 Bill` < quantile (cc$`2005-09-01 Bill`,c(.95))
         & cc$`2005-08-01 Bill` > quantile (cc$`2005-08-01 Bill`,c(.05)) & cc$`2005-08-01 Bill` < quantile (cc$`2005-08-01 Bill`,c(.95))
         & cc$`2005-07-01 Bill` > quantile (cc$`2005-07-01 Bill`,c(.05)) & cc$`2005-07-01 Bill` < quantile (cc$`2005-07-01 Bill`,c(.95))
         & cc$`2005-06-01 Bill` > quantile (cc$`2005-06-01 Bill`,c(.05)) & cc$`2005-06-01 Bill` < quantile (cc$`2005-06-01 Bill`,c(.95))
         & cc$`2005-05-01 Bill` > quantile (cc$`2005-05-01 Bill`,c(.05)) & cc$`2005-05-01 Bill` < quantile (cc$`2005-05-01 Bill`,c(.95))
         & cc$`2005-04-01 Bill` > quantile (cc$`2005-04-01 Bill`,c(.05)) & cc$`2005-04-01 Bill` < quantile (cc$`2005-04-01 Bill`,c(.95))
         & cc$`2005-09-01 Payment` > quantile (cc$`2005-09-01 Payment`,c(.05)) & cc$`2005-09-01 Payment` < quantile (cc$`2005-09-01 Payment`,c(.95))
         & cc$`2005-08-01 Payment` > quantile (cc$`2005-08-01 Payment`,c(.05)) & cc$`2005-08-01 Payment` < quantile (cc$`2005-08-01 Payment`,c(.95))
         & cc$`2005-07-01 Payment` > quantile (cc$`2005-07-01 Payment`,c(.05)) & cc$`2005-07-01 Payment` < quantile (cc$`2005-07-01 Payment`,c(.95))
         & cc$`2005-06-01 Payment` > quantile (cc$`2005-06-01 Payment`,c(.05)) & cc$`2005-06-01 Payment` < quantile (cc$`2005-06-01 Payment`,c(.95))
         & cc$`2005-05-01 Payment` > quantile (cc$`2005-05-01 Payment`,c(.05)) & cc$`2005-05-01 Payment` < quantile (cc$`2005-05-01 Payment`,c(.95))
         & cc$`2005-04-01 Payment` > quantile (cc$`2005-04-01 Payment`,c(.05)) & cc$`2005-04-01 Payment` < quantile (cc$`2005-04-01 Payment`,c(.95)))
         "& cc$`2005-04-01 Status` > quantile (cc$`2005-04-01 Status`,c(.05)) & cc$`2005-04-01 Status` < quantile (cc$`2005-04-01 Status`,c(.95))
         & cc$`2005-05-01 Status` > quantile (cc$`2005-05-01 Status`,c(.05)) & cc$`2005-05-01 Status` < quantile (cc$`2005-05-01 Status`,c(.95))
         & cc$`2005-06-01 Status` > quantile (cc$`2005-06-01 Status`,c(.05)) & cc$`2005-06-01 Status` < quantile (cc$`2005-06-01 Status`,c(.95))
         & cc$`2005-07-01 Status` > quantile (cc$`2005-07-01 Status`,c(.05)) & cc$`2005-07-01 Status` < quantile (cc$`2005-07-01 Status`,c(.95))
         & cc$`2005-08-01 Status` > quantile (cc$`2005-08-01 Status`,c(.05)) & cc$`2005-08-01 Status` < quantile (cc$`2005-08-01 Status`,c(.95))
         & cc$`2005-09-01 Status` > quantile (cc$`2005-09-01 Status`,c(.05)) & cc$`2005-09-01 Status` < quantile (cc$`2005-09-01 Status`,c(.95)))"

cc$`2005-09-01 Status`[cc$`2005-09-01 Status` < 1] <- 0
cc$`2005-08-01 Status`[cc$`2005-08-01 Status` < 1] <- 0
cc$`2005-07-01 Status`[cc$`2005-07-01 Status` < 1] <- 0
cc$`2005-06-01 Status`[cc$`2005-06-01 Status` < 1] <- 0
cc$`2005-05-01 Status`[cc$`2005-05-01 Status` < 1] <- 0
cc$`2005-04-01 Status`[cc$`2005-04-01 Status` < 1] <- 0

#Graph Colours

cols <- c("Male"= "#060047", "Female" = "#B3005E")
cols2 <- c("Married"= "#B3005E", "Single" = "#E90064", "Others" ="#060047")
cols3 <- c('Default'= "#060047", 'Did not default' = "#B3005E")
cols4 <- c("Graduate"= "#B3005E", "University" = "#E90064", "High School" ="#FF5CA3", "Others" = "#060047")

bd <- cc %>% mutate(Default = recode(Default, `1` = "Default", `0`= "Did not default" ))
bd$`2005-09-01 Status`[bd$`2005-09-01 Status` < 1] <- "0"
bd$`2005-08-01 Status`[bd$`2005-08-01 Status` < 1] <- "0"
bd$`2005-07-01 Status`[bd$`2005-07-01 Status` < 1] <- "0"
bd$`2005-06-01 Status`[bd$`2005-06-01 Status` < 1] <- "0"
bd$`2005-05-01 Status`[bd$`2005-05-01 Status` < 1] <- "0"
bd$`2005-04-01 Status`[bd$`2005-04-01 Status` < 1] <- "0"

```

```{r}
#| echo: false
#| error: false
kable(bd %>% group_by(Default) %>% summarize(Count=n()) %>% mutate(Percent = Count/sum(Count)*100))
```

```{r}
#| echo: false
#| warning: false
kable(bd %>% group_by(Default,Marriage) %>% summarize(Count=n()) %>% mutate("Group %" = Count/sum(Count)*100, "Global %" = Count/nrow(bd)*100) %>% mutate_if(is.numeric, round, 2), col.names=c("Default","Marital Status","Count","Group %","Global %"), caption = "Marital Status")
kable(bd %>% group_by(Default, Sex) %>% summarize(Count=n()) %>% mutate("Group %" = Count/sum(Count)*100, "Global %" = Count/nrow(bd)*100) %>% mutate_if(is.numeric, round, 2), caption = "Sex")
kable(bd %>% group_by(Default, Education) %>% summarize(Count=n()) %>% mutate("Group %" = Count/sum(Count)*100, "Global %" = Count/nrow(bd)*100) %>% mutate_if(is.numeric, round, 2), caption = "Education")

```

```{r}
#| echo: false
kable(bd %>%
  group_by(bd$Default) %>%
  summarize(Balance=mean(Balance), Age=mean(Age)), col.names=c("","Balance","Age"), caption = "Balance and Age")

```


| Category       | Expected Values |
|----------------|-----------------|
| Default Status | Did not default |
| Age            | 34              |
| Sex            | Female          |
| Education      | University      |
| Marital Status | Single          |
| Balance        | 147,357         |


#### Bills, Payments, and Delays

```{r}
#| echo: false

kable(bd %>%
  group_by(bd$Default) %>%
  summarize("04/2005"=mean(cc$`2005-04-01 Bill`),"05/2005"=mean(cc$`2005-05-01 Bill`), "06/2005"=mean(cc$`2005-06-01 Bill`),"07/2005"=mean(cc$`2005-07-01 Bill`),"08/2005"=mean(cc$`2005-08-01 Bill`),"09/2005"=mean(cc$`2005-09-01 Bill`)), caption = "April 2004 - September 2005 Bill", col.names=NULL)

kable(bd %>%
  group_by(bd$Default) %>%
  summarize("04/2005"=mean(cc$`2005-04-01 Payment`),"05/2005"=mean(cc$`2005-05-01 Payment`), "06/2005"=mean(cc$`2005-06-01 Payment`),"07/2005"=mean(cc$`2005-07-01 Payment`),"08/2005"=mean(cc$`2005-08-01 Payment`),"09/2005"=mean(cc$`2005-09-01 Payment`)), caption = "April 2004 - September 2005 Payment", col.names=NULL)

kable(table(bd$Default,bd$`2005-04-01 Status`), "pipe", col.names = c("Pay duly","2 month delay", "3 month delay", "4 month delay","5 month delay","6 month delay"), caption = "April 2005 Payment Delays")
kable(table(bd$Default,bd$`2005-05-01 Status`), "pipe", col.names= c("Pay duly","2 month delay", "3 month delay", "4 month delay","5 month delay"), caption = "May 2005 Payment Delays")
kable(table(bd$Default,bd$`2005-06-01 Status`), "pipe", col.names= c("Pay duly","2 month delay", "3 month delay", "4 month delay"), caption = "June 2005 Payment Delays")
kable(table(bd$Default,bd$`2005-07-01 Status`), "pipe", col.names= c("Pay duly","1 month delay","2 month delay", "3 month delay", "4 month delay"), caption = "July 2005 Payment Delays")
kable(table(bd$Default,bd$`2005-08-01 Status`), "pipe", col.names= c( "Pay duly","1 month delay","2 month delay", "3 month delay", "4 month delay"), caption = "August 2005 Payment Delays")
kable(table(bd$Default,bd$`2005-09-01 Status`), "pipe", col.names= c("Pay duly","1 month delay","2 month delay", "3 month delay", "4 month delay"), caption = "September 2005 Payment Delays")
```

#### Defaulting next month by Age and Balance

```{r}
#| echo: false
#Age*Default
ad <- cc %>% 
  group_by(Default) %>% 
  count(Age) %>% 
  mutate(Percent = n / nrow(cc) *100, Default = recode(Default, `1` = "Default", `0`= "Did not default" ))

ad$Percent <- round(ad$Percent,2)
ggplotly(ggplot(ad, aes(x = Age, y = Percent, fill = Default)) 
         + geom_bar(stat = "identity")
         + scale_fill_manual(values=cols3)
         + labs(y ="Percent %", x="Age") 
         + theme_classic() + labs(fill='') +
           theme(plot.title = element_text(hjust = 0.5), plot.margin = margin(2, 2, 2, 2, "cm")))
```

```{r}
#| echo: false
#| warning: false
#Balance by Default
ggplotly(ggplot(bd, aes(x = Balance, fill = Default)) 
         + geom_bar(aes(y = (..count..)/sum(..count..)*100), width=6000)
         + scale_fill_manual(values=cols3)
         + labs(y ="Percent %", x="Balance") + labs(fill='') 
         + theme_classic()+
           theme(plot.title = element_text(hjust = 0.5), plot.margin = margin(2, 2, 2, 2, "cm")))
```

## Predictors

### By the demographic

> Chi-Square tests

```{r include=FALSE}
#| echo: true
#| #| output: false
c1 <- chisq.test(cc$Default,cc$Sex)
c2 <- chisq.test(cc$Default,cc$Education)
c3 <- chisq.test(cc$Default,cc$Marriage)
```

##### Sex

```{r include=FALSE}
#| code-fold: true
#| echo: true
#| output: false
c1[["p.value"]]
```

|          |              |
|----------|--------------|
| P-Value: | 0.0002327589 |

Defaulting next month and Sex are related.

##### Education

```{r include=FALSE}
#| code-fold: true
#| echo: true
#| output: false
c2[["p.value"]]
```

|          |               |
|----------|---------------|
| P-Value: | 0.00007747524 |

Defaulting next month and Education are related.

##### Marital Status

```{r include=FALSE}
#| code-fold: true
#| echo: true
#| output: false
c3[["p.value"]]
```

|          |             |
|----------|-------------|
| P-Value: | 0.003104148 |

Defaulting next month and Marital Status are related.

> Logistic Regressions were determined by significance, residuals, ROC, and link function ; AIC is also provided to compare models.

```{r include=FALSE}
#| code-fold: true
#| echo: true
#| output: false

l1<-summary(glm(Default ~ Sex + relevel(factor(Education), ref="University") + relevel(factor(Marriage), ref="Single"), data = cc, family = binomial))
format(l1[["coefficients"]], scientific=TRUE)
```

```{r include=FALSE}
#| code-fold: true
#| echo: true
#| output: false
l1[["aic"]]

```

**Baseline: Female, University, and Single.**

|      |          |
|------|----------|
| AIC: | 7020.561 |

| Variable                 | Coefficient   | P-Value      |
|--------------------------|---------------|--------------|
| Sex -Male                | 3.942187e+00  | 8.074185e-05 |
| Education - Graduate     | -3.119592e+00 | 1.811016e-03 |
| Education - Others       | -2.255810e+00 | 2.408254e-02 |
| Marital Status - Married | 2.567441e+00  | 1.024522e-02 |

Compared to the baseline, the chances of defaulting next months payment are predicted to:

-   Increase with the Sex of the borrower being Male.

-   Increase with being Married.

-   Decrease with Education being Graduate or Others.

### By the numbers

```{r include=FALSE}
#| code-fold: true
#| echo: true
#| output: false

l2<-summary(glm(Default ~ Balance + `2005-09-01 Payment` + `2005-09-01 Status` + `2005-08-01 Status` + `2005-07-01 Status` +`2005-04-01 Status` , data = cc, family = binomial))
format(l2[["coefficients"]],scientific=TRUE)
```

```{r include=FALSE}
#| code-fold: true
#| echo: true
#| output: false
l2[["aic"]]
```

|      |          |
|------|----------|
| AIC: | 5956.876 |

| Variables          | Coefficients  | P-Value       |
|--------------------|---------------|---------------|
| Balance            | -2.568599e-06 | 4.732495e-10  |
| 2005-09-01 Payment | -1.001972e-04 | 7.931939e-10  |
| 2005-09-01 Delay   | 1.388836e+00  | 9.430681e-117 |
| 2005-08-01 Delay   | -3.579733e-01 | 3.093678e-04  |
| 2005-07-01 Delay   | 4.012861e-01  | 4.681578e-05  |
| 2005-04-01 Delay   | 2.252740e-01  | 1.759120e-02  |

The chances of defaulting next months payment are predicted to:

-   Decrease with a greater credit balance.

-   Decrease with a greater most recent payment.

-   Increase with delay in payment for September, July and April. Surprisingly, it decreases with delay in payment in August.

Age is not a significant predictor.

------------------------------------------------------------------------

### In Concert

```{r include=FALSE}
#| code-fold: true
#| echo: true
#| output: false
l3<-summary(glm(Default ~ Balance + Sex + relevel(factor(Education), ref="University") + relevel(factor(Marriage), ref="Single") + `2005-09-01 Payment` + `2005-09-01 Status` + `2005-08-01 Status` + `2005-07-01 Status` +`2005-04-01 Status`, data = cc, family = binomial))
format(l3[["coefficients"]],scientific=TRUE)
```

```{r include=FALSE}
#| code-fold: true
#| echo: true
#| output: false
l3[["aic"]]
```

**Baseline: Female, University, and Single.**

|     |          |
|-----|----------|
| AIC | 5936.432 |

| Variables                | Coefficients  | P-Value       |
|--------------------------|---------------|---------------|
| Balance                  | -5.753312e+00 | 8.751180e-09  |
| 2005-09-01 Payment       | -5.961952e+00 | 2.492422e-09  |
| 2005-09-01 Delay         | 2.304838e+01  | 1.527069e-117 |
| 2005-08-01 Delay         | -3.734682e+00 | 1.879528e-04  |
| 2005-07-01 Delay         | 4.085407e+00  | 4.399971e-05  |
| 2005-04-01 Delay         | 2.350145e+00  | 1.876611e-02  |
| Sex - Male               | 3.750723e+00  | 1.763254e-04  |
| Education - Others       | -2.217312e+00 | 2.660177e-02  |
| Marital Status - Married | 3.212293e+00  | 1.316799e-03  |

The chances of defaulting next months payment are significantly predicted to:

-   Decrease with a greater credit balance.

-   Decrease with a greater most recent payment.

-   Increase with payment delay status in September, July and April. Like before, it decreases with payment delay status in August.

-   Increase when the Sex is Male.

-   Increase with being Married.

-   Decrease Education being Others

## Conclusion

Based on the data, a higher credit balance and a higher payment made towards the previous month's bill are associated with a lower chance of defaulting on a credit account in the next month.

While delays in payment are significant in predicting default for September, July, and April; August predicts no default in October.

Compared to the expected borrower of Female, University educated, and Single: Male borrowers increased defaulting chances, Married borrowers increased defaulting chances, and graduates that considered themselves "Other" decreased defaulting chances.
